\chapter{Evaluation of the Productivity Checker}
\label{cha:evaluation}

In Section~\ref{sec:less-restr-prod}, we discussed the motivation behind
implementing a less restrictive productivity checker. In
Chapter~\ref{cha:infer-guard-recurs}, we introduced a system for
productivity checking functions by inferring guarded recursive definitions from
user-written definitions. In the following, we evaluate our implementation of
this system. We investigate both the kinds of functions that can now be automatically proven total by our
system, and those that cannot. The
implementation of all functions discussed in this section can be found in
Appendix~\ref{app:example-programs}. 

We divide productive functions into three classes: 
\begin{itemize}
\item Functions we can prove total.
\item Functions we should be able to prove total, but can not.
\item Functions we should not and can not prove total.
\end{itemize}

The first class is self-explanatory. The second stems from our implementation. While
we do indent to fully implement the system described in
Chapter~\ref{cha:infer-guard-recurs}, our current implementation has not quite
reached full coverage yet. The last covers functions that are productive, but that we
can not infer a guarded recursive definition for due to scoping restrictions.
% Intro with link back to motivation.
% Three classes of functions:
% What we can check
% What we should be able to check
% What we should not be able to check.
% Why can we check what we can?
% Why can't we check what we should be able to?
% Why should we not be able to check what we shouldn't?
% Modality

Figure~\ref{fig:productivity_table_1} shows examples of functions falling into
the first category. While some functions can already be proven productive by the
current productivity checker in Idris which uses the syntactic guardedness
principle, we can see that we have added to the set of Idris functions
automatically proven productive. We can, for example, now prove functions
productive where the recursive reference is given as an argument to another
function.

\begin{figure}[h]
\begin{center}
  \begin{tabular}{| l | c | c |} \hline
    Function & Syntactic Guardedness & Guarded Recursion \\ \hline
    \texttt{zeros} & Productive & Productive \\ \hline
    \texttt{repeat} & Productive & Productive \\ \hline
    \texttt{map} & Productive & Productive \\ \hline
    \texttt{zipWith} & Productive & Productive \\ \hline
    \texttt{toggle} & Productive & Productive \\ \hline
    \texttt{interleave} & Productive & Productive \\ \hline
    \texttt{unfold} & Productive & Productive \\ \hline
    \texttt{prepend} & Productive & Productive \\ \hline
    \texttt{cycle} & Productive & Productive \\ \hline
    \texttt{nats} & Not Productive & Productive \\ \hline
    \texttt{fib} & Not Productive & Productive \\ \hline
    \texttt{fac} & Not Productive & Productive \\ \hline
    \texttt{paperfold} & Not Productive & Productive \\ \hline
    \texttt{calculateWilfully} & Productive & Productive \\ \hline
    \texttt{tmap} & Productive & Productive \\ \hline
    \texttt{tzip} & Productive & Productive \\ \hline
    \texttt{trepeat} & Productive & Productive \\ \hline
    \texttt{carry} & Not Productive & Productive \\ \hline
    \texttt{pingserver'} & Not Productive & Productive \\ \hline
    \texttt{imap} & Productive & Productive \\ \hline
    \texttt{counter} & Not Productive & Productive \\ \hline
  \end{tabular}
\end{center}
  \caption{Examples of functions we can prove productive using our guarded
    recursive system. The header ``Syntactic Guardedness'' refers to the current
    implementation of the syntactic guardedness principle, and ``Guarded Recursion''
    refers to our implementation of guarded recursion in Idris.}
  \label{fig:productivity_table_1}
\end{figure}

Figure~\ref{fig:productivity_table_2} shows examples of functions falling into
the second category. These are functions for which our implementation can not
yet infer a guarded term, although it should be able to in theory. As such, a better
implementation would most likely show that these are, in fact, productive. We
will evaluate our implementation further in Section~\ref{sec:our-implementation}.

\begin{figure}[h]
\begin{center}
  \begin{tabular}{| l | c | c |} \hline
    Function & Syntactic Guardedness & Guarded Recursion \\ \hline
    \texttt{never} & Not Productive & Not Productive \\ \hline
    \texttt{bind} & Not Productive & Not Productive \\ \hline
    \texttt{multmachine} & Not Productive & Not Productive \\ \hline
    \texttt{pingserver} & Not Productive & Not Productive \\ \hline
  \end{tabular}
\end{center}
  \caption{Examples of functions we can not yet prove productive. Productivity
    of these
    functions should be provable within our system.}
  \label{fig:productivity_table_2}
\end{figure}

The last class of functions are functions that we do not expect our system to be
able to prove productive, not even in theory. While only a few examples are listed in
Figure~\ref{fig:productivity_table_3}, we expect that there are many more of such
definitions. The problem with these functions is that they are neither causal
nor non-causal, and therefore require a very specific guarded recursive type to be proven productive.

\section{Function Types}
\label{sec:function-types}
To simplify the inference of guarded recursion, we chose to classify all
function definitions as either causal or non-causal. We used this to help us infer the guarded type of a function. There
are, however, functions for which the guarded type falls outside this
classification. The examples in Figure~\ref{fig:productivity_table_3} both need
more sophisticated types to be proven total.

\begin{figure}[h]
\begin{center}
  \begin{tabular}{| l | c | c |} \hline
    Function & Syntactic Guardedness & Guarded Recursion \\ \hline
    \texttt{mergef} & Not Productive & Not Productive \\ \hline
    \texttt{paperfold'} & Not Productive & Not Productive \\ \hline
  \end{tabular}
\end{center}
  \caption{Examples of functions that we can not prove productive. Productivity
    of these functions should not currently be provable within our system.}
  \label{fig:productivity_table_3}
\end{figure}

Atkey and McBride\,\citep{Atkey:2013} prove \texttt{mergef} productive by
changing the type of the input function. They alter it to be:

\begin{lstlisting}[mathescape, title=\idrisBlock]
mergef : (Nat $\to$ Nat $\to \laterkappan$Stream$^\kappa$ Nat $\to$ Stream$^\kappa$) $\to$
         Stream$^\kappa$ Nat $\to$ Stream$^\kappa$ Nat $\to$ Stream$^\kappa$ Nat
mergef f s t = f (head s) (head t) 
                 (mergef $\tensor^\kappa_1$ (Next$^\kappa$ f) $\tensor^{\kappa}_1$ (tail s) $\tensor^\kappa_1$ (tail t))
\end{lstlisting}

This follows the intuition that the \texttt{Stream$^\kappa$} argument to \texttt{f} must be used in a
later context. It allows us to give the recursive reference, which is always
later, as an argument to \texttt{f}.

In a similar fashion, Clouston et al.\,\citep{BirkedalL:guarded-lambda-conf}
prove \texttt{paperfold'} productive by using a differently typed \texttt{interleave}:

\begin{lstlisting}[mathescape, title=\idrisBlock]
interleave : Stream$^\kappa$ Nat $\to$ $\laterkappan$ Stream$^\kappa$ Nat $\to$ Stream$^\kappa$ Nat
\end{lstlisting}

This type captures the intuition that while the first argument is needed straightaway, the
second argument is always only needed later. With this type for interleave, \texttt{paperfolds'} can be defined by guarded recursion as
follows:

\begin{lstlisting}[mathescape, title=\idrisBlock]
paperfolds' : Stream$^\kappa$ Nat
paperfolds' = interleave toggle paperfolds'
\end{lstlisting}

Inferring these types would be no easy task, and has been deemed out of scope for
this project. With better support for user interaction with the
inference system, one could imagine a system where the user could specify these
types manually. The inference system would then still be
inferring the term, but not the type.

% \begin{figure}[h]
% \begin{lstlisting}[mathescape]
% semiinterleave : $\forall \kappa$.Stream$^\kappa$ a $\to$ $\forall \kappa$(Stream$^\kappa$ a $\to$ Stream$^\kappa$ a)
% semiinterleave s = $\Lambda \kappa.$ semiinterleave'(s)

% semiinterleave' : $\forall \kappa$.Stream$^\kappa$ a $\to$ Stream$^\kappa$ a $\to$ Stream$^\kappa$ a
% semiinterleave' = pfix(f)

% f : ($\forall \kappa$.Stream$^\kappa$ a $\to$ $\later$$^{\kappa}$(Stream$^\kappa$ a $\to$ Stream$^\kappa$ a)) $\to$ 
%        $\forall \kappa$.Stream$^\kappa$ a $\to$ Stream$^\kappa$ a $\to$ Stream$^{\kappa}$ a
% f g xs s = StreamCons (head xs)
%     (Next (StreamCons (head s) 
%                       ((g (tail (tail xs))) $\tensor^\kappa$(tail s))))
% \end{lstlisting}
% \caption{A function interleave a stream with the even indices of another
%   stream.}
% \label{fig:semiinterleave}
% \end{figure}

\section{Our Implementation}
\label{sec:our-implementation}
As discussed, the implementation of our system is not yet complete. There are
definitions that we should be able to check, but can not. The cause of this is
simply a flawed implementation. In the later parts of the project we have
prioritized fine-tuning our system on a theoretical level, rather than ironing
out bugs in our implementation. This has resulted in a stronger system, but a
weaker implementation.

Our implementation, while flawed, demonstrates that implementing the system
described in Chapter~\ref{cha:infer-guard-recurs} is feasible. It shows that,
through a set of inference rules, it is possible to infer a guarded recursive TT
term from a given input TT term. This term is then checked against the guarded
recursion rules to ensure productivity.

We are convinced that it is possible to improve our implementation, such that it
resembles the system presented in Chapter~\ref{cha:infer-guard-recurs} more
closely. There are improvements to be made, but what we have done
thus far has already widened the range of function that the Idris compiler can
prove productive, allowing for
more total definitions to be written.
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../copatterns-thesis"
%%% End:
